---
title: "Draft paper"
author: "Nick Sunderland"
date: "2024-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dir <- here::here()
library(data.table)
library(ggplot2)
library(scales)
library(plotly)
library(reticulate)
reticulate::py_run_string("import plotly")
reticulate::py_run_string("import kaleido")
library(viridis)
library(splines)
library(kableExtra)
```

## Set default parameters
```{r}
dcm_prev      <- 0.004 # DCM prevalence
gene_pos_pct  <- 0.35  # proportion of DCM cases gene positive
gene_vus_pct  <- 0.15  # proportion of DCM cases variant of unknown significance
gene_neg_pct  <- 0.50  # proportion of DCM cases gene negative
n_individuals <- 500000
```

## Simulate data
Here we simulate a dataset of `r format(n_individuals, big.mark = ",", scientific=FALSE)` individuals using a DCM prevalence of `r sprintf("%0.2f%%", dcm_prev*100)` and percentage of gene positive, gene negative, and variant of unknown significance (VUS) of `r sprintf("%.0f%%", gene_pos_pct*100)`, `r sprintf("%.0f%%", gene_neg_pct*100)`, and `r sprintf("%.0f%%", gene_vus_pct*100)`, respectively. We adjust the means of the polygenic risk score such that the gene positive cohort has the maximum, followed by VUS, then gene negative. Non-DCM individuals have a mean PGS of 0.

```{r, warning=FALSE, message=FALSE}
# set prevalences and PGS means
gene_statuses <- list(
  `No DCM`       = list(prop = 1-dcm_prev,            mean_pgs = 0.00),  
  `DCM gene +ve` = list(prop = dcm_prev*gene_pos_pct, mean_pgs = 1.00), 
  `DCM gene vus` = list(prop = dcm_prev*gene_vus_pct, mean_pgs = 0.85), 
  `DCM gene -ve` = list(prop = dcm_prev*gene_neg_pct, mean_pgs = 0.60)   
)

# reproducibility
set.seed(123)

# simulate te dataset
dat <- data.table(
  id          = sprintf(paste0("ID_%0", ceiling(log10(n_individuals+1)), "d"), 1:n_individuals), 
  gene_status = factor(sample(x       = names(gene_statuses), 
                              size    = n_individuals, 
                              replace = TRUE, 
                              prob    = sapply(gene_statuses, function(x) x$prop)), 
                      levels = names(gene_statuses))
)
dat[, dcm_status := factor(gene_status=="No DCM", levels=c(T,F), labels=c("No DCM", "DCM"))]
dat[, pgs := rnorm(.N, mean = gene_statuses[[as.character(gene_status)]]$mean_pgs, sd = 1), by = .(gene_status)]

# view
knitr::kable(unique(dat, by="gene_status"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.show="hold", fig.align='center', out.width="50%"}
# plot pie chart
p <- ggplot(data = (dat[gene_status != "No DCM", .N, by = gene_status
                       ][
                        , prop := N / sum(N)
                       ]), 
       aes(x = "", y = prop, fill = gene_status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  scale_fill_viridis_d() +
  labs(fill = "Gene status") +
  theme_void(base_size = 22)
p
png(file.path(dir, "figures", "dcm_proportions.png"), width=600, height=600)
p
invisible(dev.off())

# plot PGS distributions
p1 <- ggplot(dat, aes(x = pgs, fill = gene_status)) +
  geom_density(alpha=0.3) +
  scale_fill_viridis_d() +
  labs(x = "PGS", fill = "Gene status") +
  theme_classic(base_size = 22) +
  theme(axis.text.x  = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.y  = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank())
p1
png(file.path(dir, "figures", "psg_distributions.png"), width=800, height=600)
p1
invisible(dev.off())
```

---

## Odds ratio by DCM PGS quantile
We then calculate the odds ratio for DCM by PGS bin. We define appropriate cut-offs for the PGS for "pathogenic" and "intermediate" criteria.

### PGS odds ratio criteria
```{r, echo=F}
pgs_thresholds <- data.table(
  or_thresh  = c(1.5, 5.0), 
  thresh_lab = c("intermediate", "pathogenic") 
)
knitr::kable(pgs_thresholds) |> 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Odds ratio 
We calculate the odds ratio by PGS bin and fit a spline model to help extract the quantile at which the thresholds are reached.

```{r, warning=FALSE, message=FALSE}
# calculate the odds ratio of DCM by PGS quantile 
n_quantiles <- 10
quantiles   <- c(0, 1, 5, 10, 20, 40, 60, 80, 90, 95, 99, 100)
dat[, pgs_quantile := cut(pgs, 
                          breaks = quantile(pgs, probs = quantiles/100, na.rm = TRUE), 
                          labels = c(paste0("<",quantiles[2]), 
                                     paste0(quantiles[2:(length(quantiles)-2)], "-", quantiles[3:(length(quantiles)-1)]), 
                                     paste0(">", quantiles[length(quantiles)-1])), 
                          include.lowest = TRUE)]
dat[, pgs_percentile := cut(pgs, 
                            breaks = quantile(pgs, probs = seq(0,1,0.01), na.rm = TRUE), 
                            labels = seq(1,100,1), 
                            include.lowest = TRUE)]
or_table <- data.table(pgs_quantile = unique(dat$pgs_quantile))
or_table[, `:=`(
  dcm_cases_in_group    = dat[pgs_quantile == .BY[[1]] & dcm_status != "No DCM", .N],
  dcm_controls_in_group = dat[pgs_quantile == .BY[[1]] & dcm_status == "No DCM", .N],
  dcm_cases_rest        = dat[pgs_quantile != .BY[[1]] & dcm_status != "No DCM", .N],
  dcm_controls_rest     = dat[pgs_quantile != .BY[[1]] & dcm_status == "No DCM", .N]
), by = pgs_quantile]
or_table[, `:=`(
  odds_in_group = dcm_cases_in_group / dcm_controls_in_group,
  odds_in_rest  = dcm_cases_rest / dcm_controls_rest
)]
or_table[, `:=`(odds_ratio    = odds_in_group / odds_in_rest, 
                se_log_or     = sqrt(1 / dcm_cases_in_group + 
                                     1 / dcm_controls_in_group + 
                                     1 / dcm_cases_rest + 
                                     1 / dcm_controls_rest))]
or_table[, `:=`(log_or   = log(odds_ratio))]
or_table[, `:=`(ci_lower = exp(log_or - 1.96 * se_log_or),
                ci_upper = exp(log_or + 1.96 * se_log_or))]

# model the odds ratio across PGS quantile and extract the thresholds of 5 and 1.5
or_table[, pgs_midpoint := as.numeric(sub("-.*", "", as.character(pgs_quantile))) + (as.numeric(sub(".*-", "", as.character(pgs_quantile))) - as.numeric(sub("-.*", "", as.character(pgs_quantile)))) / 2]
or_table[as.character(pgs_quantile) == "<1", pgs_midpoint := 0.5]
or_table[as.character(pgs_quantile) == ">99", pgs_midpoint := 99.5]

spline_model <- lm(log(odds_ratio) ~ ns(pgs_midpoint, df = 6), data = or_table)
pred_pgs_values <- seq(0,100,0.01)
predicted_or <- data.table(pgs_quantile_val = pred_pgs_values, 
                           pred_or      = exp(predict(spline_model, newdata = data.frame(pgs_midpoint = pred_pgs_values))))

# predict the thresholds from the fit 
pgs_thresholds[, pgs_cutoff := sapply(or_thresh, function(x) predicted_or[which.min(abs(pred_or-x)), pgs_quantile_val])]
pgs_thresholds[, pgs := sapply(pgs_cutoff, function(x) dat[ceiling(x)==as.numeric(pgs_percentile), mean(pgs, na.rm=T)])]

dat <- dat[
  pgs_thresholds, 
  on = .(pgs > pgs), 
  `:=`(thresh_lab = paste0(i.thresh_lab, " PGS"))
]
dat[is.na(thresh_lab), `:=`(thresh_lab = "low PGS")]

# view
knitr::kable(unique(dat, by="thresh_lab"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', out.width="49%"}
# plot OR of DCM by PGS quantile 
p3 <- ggplot(or_table, aes(x = pgs_midpoint, y = odds_ratio, ymin = ci_lower, ymax = ci_upper)) +
  geom_hline(yintercept = 1.0, color="darkgray", linetype = "dotted") +
  geom_errorbar(width=0.3, color="darkgray") + 
  geom_point(size=4) +
  geom_line(data = predicted_or, 
            aes(x = pgs_quantile_val, y = pred_or, group=1), color = "darkred", inherit.aes = FALSE) +
  geom_segment(data = pgs_thresholds, aes(x=0, xend = pgs_cutoff, y = or_thresh, yend = or_thresh, color = thresh_lab), inherit.aes = F, linetype = "dashed") +
  geom_segment(data = pgs_thresholds, aes(y=0, yend = or_thresh, x =pgs_cutoff, xend=pgs_cutoff, color = thresh_lab), inherit.aes = F, linetype = "dashed") +
  geom_text(data = pgs_thresholds, aes(x=0, y = or_thresh, label = paste0(thresh_lab, " OR=",or_thresh, " (", sprintf("%.0f%%", pgs_cutoff), ")"),  color = thresh_lab), vjust=-1, hjust=0, inherit.aes = F, show.legend = FALSE) +
  theme_classic(base_size = 22) +
  labs(x = "PGS quantile", 
       y = "OR for DCM") +
  guides(color = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=1))
p3
png(file.path(dir, "figures", "odds_ratio_psg.png"), width=600, height=600)
p3
invisible(dev.off())

# plot the distributions of PGS in affected and unaffected
p2 <- ggplot(dat, aes(x = pgs, fill = dcm_status)) +
  geom_density(alpha=0.3) +
  geom_vline(data = pgs_thresholds, aes(xintercept = pgs), color="red") +
  geom_text(data = pgs_thresholds, aes(x=pgs, y = 0, label = paste0(thresh_lab, " OR=",or_thresh)), size=6, color = "red", vjust=-0.5, hjust=0, angle=90, inherit.aes = F, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(x = "PGS", fill = "DCM status") +
  theme_classic(base_size = 22) +
  theme(axis.text.x  = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.y  = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank())
p2
png(file.path(dir, "figures", "psg_dcm_distributions.png"), width=800, height=600)
p2
invisible(dev.off())
```

### Odds ratio thresholding and increase in diagnostic yeild
We then annotate the DCM cases in the dataset as to whether they have a pathogenic, intermediate or low PGS. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', out.width="50%"}
# plot the pie chart of DCM gene status by PGS thresholds
sunburst_data <- rbind(
  dat[dcm_status == "DCM", .(gene_status = "DCM", thresh_lab = "", N = .N)],
  dat[dcm_status == "DCM", .(thresh_lab = "", N = .N), by = "gene_status"],
  dat[dcm_status == "DCM", .N, by = .(gene_status, thresh_lab)]
)[, `:=`(id     = trimws(paste(gene_status, thresh_lab)),
         parent = fcase(gene_status=="DCM", "", 
                        thresh_lab==""    , "DCM", 
                        thresh_lab!=""    , as.character(gene_status)))]
sunburst_data[, label := fcase(gene_status=="DCM", "DCM", 
                               thresh_lab==""    , as.character(gene_status),
                               thresh_lab!=""    , thresh_lab)]
sunburst_data <- sunburst_data[, .(id, parent, N, label)]

# Create the sunburst plot
p4 <- plot_ly(
  sunburst_data,
  ids     = ~id,
  labels  = ~label,
  parents = ~parent,
  values  = ~N,
  type    = 'sunburst',
  text = ~N,
  branchvalues = 'total'
) |> layout(colorway = c("#345799", "#e6550d", "#9ecae1"))
p4
save_image(p4, file.path(dir, "figures", "gene_status_by_psg_threshold.png"))
```


## Likelihood ratios and population attributable risk
We derive the odds-ratio per standard deviation chnage in PGS for DCM use this to calculate the likelihood ratios, and population attributable risk, by PGS bin.

```{r}
# Likelihood ratios and attributable risk faction
# calculate the odds ratio per Sdev of the PGS
model             <- glm(dcm_status ~ pgs, data = dat, family = "binomial")
summary_model     <- summary(model)
odds_ratio_sd     <- exp(coef(summary_model)["pgs", "Estimate"] * sd(dat$pgs))
odds_ratio_sd_lci <- exp((coef(summary_model)["pgs", "Estimate"] - 1.96*coef(summary_model)["pgs", "Std. Error"]) * sd(dat$pgs))
odds_ratio_sd_uci <- exp((coef(summary_model)["pgs", "Estimate"] + 1.96*coef(summary_model)["pgs", "Std. Error"]) * sd(dat$pgs))

attributable_risk_frac <- function(prev, or_per_sd, centile) {
  odds   <- prev / (1-prev)
  u.A    <- log(or_per_sd)
  u.U    <- 0
  u.diff <- u.A - u.U
  z.U    <- qnorm(centile)
  z.A    <- z.U - u.A
  pgs_centile_aff <- pnorm(z.A)
  L.U    <- dnorm(z.U)
  L.A    <- dnorm(z.A)
  LR     <- L.A / L.U
  posttest.odds <- odds * LR
  posttest.P    <- posttest.odds / (posttest.odds + 1)
  afe           <- 1 - (prev / posttest.P)
  list(LR = LR, afe = afe)
}

likelihood_ratio <- data.table(
  dcm.P             = dcm_prev, 
  pgs_or_per_sd     = odds_ratio_sd,
  pgs_or_per_sd_lci = odds_ratio_sd_lci,
  pgs_or_per_sd_uci = odds_ratio_sd_uci,
  pgs_centile_unaff = seq(0.01,0.99,0.01)
)
likelihood_ratio[, c("LR", "afe")         := attributable_risk_frac(dcm.P, pgs_or_per_sd, pgs_centile_unaff)]
likelihood_ratio[, c("LR_lci", "afe_lci") := attributable_risk_frac(dcm.P, pgs_or_per_sd_lci, pgs_centile_unaff)]
likelihood_ratio[, c("LR_uci", "afe_uci") := attributable_risk_frac(dcm.P, pgs_or_per_sd_uci, pgs_centile_unaff)]
```


```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', out.width="50%"}
# Create the PGS centile vs proportion of DCM cases plot 
sd_lines <- data.table(y = c(ecdf(dat$pgs)(mean(dat$pgs) - 2*sd(dat$pgs)),
                             ecdf(dat$pgs)(mean(dat$pgs) - sd(dat$pgs)),
                             ecdf(dat$pgs)(mean(dat$pgs)),
                             ecdf(dat$pgs)(mean(dat$pgs) + sd(dat$pgs)),
                             ecdf(dat$pgs)(mean(dat$pgs) + 2*sd(dat$pgs))), 
                       x = 1,
                       text  = c("-2SD", "-1SD", "Mean", "+1SD", "+2SD"))

p5 <- ggplot((dat[, .(dcm_count = sum(dcm_status == "DCM"), total_count = .N), by = pgs_percentile]
           [order(as.numeric(pgs_percentile)), ]
           [, `:=`(cumulative_dcm_count   = cumsum(dcm_count))]
           [, `:=`(cumulative_proportion  = cumulative_dcm_count / dat[dcm_status == "DCM", .N])]
           [, `:=`(se                     = sqrt((cumulative_proportion * (1 - cumulative_proportion)) / dat[dcm_status == "DCM", .N]))]), 
       aes(y = as.numeric(pgs_percentile)/100, x = cumulative_proportion, group=1)) +
  geom_hline(data = sd_lines, aes(yintercept=y), linetype = "dotted", linewidth=1) +
  geom_text(data = sd_lines, aes(y=y, x=x, label=text), vjust=-0.3, hjust=0, size=5) +
  geom_line() + 
  geom_ribbon(aes(xmin = cumulative_proportion - 1.96*se, xmax = cumulative_proportion + 1.96*se), alpha = 0.2) + 
  labs(y = "PGS Percentile", x = "Proportion of DCM Cases") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), labels = percent) + 
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), labels = percent, expand = c(0,0.1)) + 
  theme_minimal(base_size = 18) +
  theme(panel.grid.minor = element_blank())
p5
png(file.path(dir, "figures", "pgs_by_pct_dcm_cases.png"), width=600, height=500)
p5
invisible(dev.off())


# plot the likelihood ratio by PGS centile
p6 <- ggplot(likelihood_ratio, aes(x = pgs_centile_unaff, y = LR)) +
  geom_line(color = "darkgray") +
  geom_point(size=3) +
  scale_x_continuous(breaks = seq(0,1,0.1), labels = percent) +
  labs(x = "PGS centile in the unaffected", y = "Likelihood Ratio") +
  theme_classic(base_size = 18)
p6
png(file.path(dir, "figures", "lr_by_pgs.png"), width=600, height=500)
p6
invisible(dev.off())

# plot the attributable risk factor for any given PGS centile
p7 <- ggplot(likelihood_ratio[afe>=0], aes(x = pgs_centile_unaff, y = afe)) +
  geom_ribbon(aes(ymin=afe_lci, ymax=afe_uci), alpha = 0.3) +
  geom_point(size=3) +
  scale_y_continuous(breaks = seq(0,1,0.1), labels = percent) +
  scale_x_continuous(breaks = seq(0,1,0.05), labels = percent) +
  labs(x = "PGS centile", y = "Attributable fraction among DCM") +
  theme_classic(base_size = 18)
p7
png(file.path(dir, "figures", "afe_by_pgs.png"), width=600, height=500)
p7
invisible(dev.off())
```



## Introduction

Y...

---

## Methods

Y...

---

## Results

Y...

---

## Discussion 

Y...

---

## References
